---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%",
  fig.path = "docs/README-"
)
```
# hellomojo

An Example R Package that Uses 'mojo' Through the `.Call` Interface 

## Overview

This package provides a setup to call [Mojo](https://www.modular.com/mojo) code from R using the mojo shared library build mode. This was obviously designed for python interop. If we only use simple types, we can call mojo using the `.Call` interface. Some boilerplate is required, but this can be automated.

## How It Works ?

Mojo code is compiled to a shared library, in our case `libhello.so`. A C file [`RC_hellomojo.c`](src/RC_hellomojo.c) provides wrappers and registration so R can call the Mojo functions via the `.Call` interface. R functions like `hellomojo()` call the native code using `.Call()`(can t see the hello world in the Readme bc it is written in C stdout). The [`configure`](configure) script uses [pixi](https://pixi.sh/) to install mojo and run the shared library build.

## Example

```{r}
# Load the package and call the native function
hellomojo::hellomojo()
hellomojo::hellomojo_add(10, 30)
```

the mojo code

```{r engine='bash', comment=''}
cat inst/mojo/hellomojo/hellomojo.mojo
```

the C wrappers

```{r engine='bash', comment=''}
cat src/RC_hellomojo.c
```

## Limitations to investigate

Of course here we are using pixi to get Mojo binaries, installing 
this on the R windows toolchain is not given. We should use uv or make a R package that install pix. Moreover i could not add windown to the current pixi workspace. This package should work fine on unix. Moreover we are calling the mojo shared object in the same address space as R even though the toolchains were different.

Additionally we have an additional C wrapping which may be not required if we pass the data directly to the mojo C callables.

## References

[Mojo Getting Started Guide](https://docs.modular.com/mojo/manual/get-started)  
[pixi: Package and Environment Manager](https://pixi.sh/)

[simpleCall example](https://github.com/coolbutuseless/simpleCall)